Gradual Soundness: Lessons from Static Python

### 2022-01-10 pj-7.1/paper.tex
pj: sec 4 complete draft, python + opts

\section{Formal Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

@@

    \inferrule*[left=Matr]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\
      \spteval_1 \spmatr \spteval_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

@@

%% TODO discuss optimizations here, not in sec 2

@@

%% TODO wherever we discuss optimizations, also talk about de-slowdowns
%% Carl: not sure where this belongs exactly, but we go to a lot of work in the
%%   compiler to intelligently minimize runtime checks by only inserting them when
%%   the source value is of dynamic type (this includes also skipping CHECK_ARGS if
%%   the function was statically invoked by another static function and thus we
%%   know the compiler checked argument types)

[FILL]

Purpose: explain significant gaps between the model and the implementation.
E.g. what would the TypeScript team need to know before using our model
to add soundness?

\subsection{Interactions with Python}

FILL

model with Python ... two languages ... boundaries ... argue for soundness

%% C-types are among the most complex because of casts? no not really complex, perhaps notable

@@

By convention, a \code{CHECK\_ARGS} instruction always appears on the first
line of a typed function body.
The optimizer uses this convention to skip argument checks by jumping
past them when it is safe to do so; namely, whenever a typed function (or method)
gets called with well-typed arguments.
%% TODO ask C&D about materialize for fn calls

@@

>>>>

\section{Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

@@

    \inferrule*[left=Matr]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\
      \spteval_1 \spmatr \spteval_0
      %% side condition: result type is not a primitive
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

@@

@@

This section explains significant aspects of \SP{} that lie outside the
scope of our model.
These aspects include sound interactions with Python code,
the bytecode language and optimizations, and dynamic Python features
that \SP{} does not yet ascribe types to.

\subsection{Open-World Python Interactions}

Although \SP{} is technically a new language, it is designed for gradual adoption.
Python programmers should be able to add types one module at a time to an
existing codebase.
Consequently, \SP{} supports interactions with Python modules in the only viable
way: by letting the Python code run with minimal constraints.
After all, if programmers need to fix every module in a codebase before they can
run tests after typing a first module, then the gradual story is lost.

Interactions with open-world~\cite{vss-popl-2017} Python code pose a minor
threat to soundness.
Typed functions (and methods) cannot assume that their arguments are
well-typed.
Arguments from typed contexts can be validated by either static checks
or materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
\Cref{s:optimize} explains how the optimizer avoids some of these checks.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else
%% ... I guess the trick is that primitives don't materialize in typed code

FILL sound generics / checked types also need to guard their writes

There are two cases, however, where new \SP{} types do add constraints to
existing Python code.
First, sound generic types (e.g. $\sptchkdict{\sptype_0}{\sptype_1}$)
force value-suppliers to use a checked constructor.
Existing code can \emph{use}\/ a checked generic via the same API as on
un-checked one, but can create new values only using the \SP{} constructors.
Second, the strict setting for modules prevents updates to module-level variables.
If existing code happens to mutate a member of a newly-strict module, a run-time
error will occur.

@@

@@

\paragraph{Indirect Optimization}

In addition to upgrading bytecode instructions to optimized ones, \SP{} takes care
to minimize the type checks that it executes at run-time.
In other words, the goal is to slow code down as little as possible.

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialze rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\ARGS} instruction
that casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function (or method) gets
called in a typed context.
In this way, all untyped calls get the necessary argument check and all typed
calls benefit from materialize checks at the call-site.
